Notes on cement modeling
========================================================

Here are the best regression models we've come up with:

```{r loaddata}
source("modelfits.R")
source("cluster.R")
```
The earth model with the basic parameters and degree 1 seems to be our best contender so far.  Here's the scatterplot:

```{r scatterplot.earth.basic.d1}
scatterplot.model(basic.earth.d1, datasets.trailing, sz=2)
```

The cluster results are also intriguing, but we haven't had a chance to do cross-validation on them yet.  Clustering on the trailing data with 5 clusters seems to produce the best results so far (but we haven't really looked at the 6-cluster data yet.)
```{r cluster5}
cmodel.trailing.5.d1 <- cluster.modelfit(km.trailing.5, complete.trailing, formula=f.basic.d1, fitfun=earth, degree=1)
cmodel.trailing.5.d2 <- cluster.modelfit(km.trailing.5, complete.trailing, formula=f.basic.d1, fitfun=earth, degree=2)
cat("cluster basic with trailing data (degree=1)\n")
cluster.rms.eval(km.trailing.5, complete.trailing, cmodel.trailing.5.d1, prn=FALSE)
cat("cluster basic with trailing data (degree=2)\n")
cluster.rms.eval(km.trailing.5, complete.trailing, cmodel.trailing.5.d2, prn=FALSE)
```

The degree 2 looks slightly better, though that's without any cross-validation applied.  Here's the scatterplot:
```{r cluster.scatterplot}
cluster.scatterplot.model(km.trailing.5, complete.trailing, cmodel.trailing.5.d2, sz=2)
```
Ugh. The results change a bit every time because of the random starting condition in k-means.  Bleah.