Notes on cement modeling
========================================================

Here are the best regression models we've come up with:

```{r loaddata}
source("modelfits.R")
source("cluster.R")
```
The earth model with the basic parameters and degree 1 seems to be our best contender so far.  Here's the scatterplot:

```{r scatterplot.earth.basic.d1}
scatterplot.model(basic.earth.d1, datasets.trailing, sz=2)
```

The cluster results are also intriguing, but we haven't had a chance to do cross-validation on them yet.  Clustering on the trailing data with 5 clusters seems to produce the best results so far (but we haven't really looked at the 6-cluster data yet.)
```{r cluster5}
cmodel.trailing.5.d1 <- cluster.modelfit(km.trailing.5, complete.trailing, formula=f.basic.d1, fitfun=earth, degree=1)
cmodel.trailing.5.d2 <- cluster.modelfit(km.trailing.5, complete.trailing, formula=f.basic.d1, fitfun=earth, degree=2)
cat("cluster basic with trailing data (degree=1)\n")
cluster.rms.eval(km.trailing.5, complete.trailing, cmodel.trailing.5.d1, prn=FALSE)
cat("cluster basic with trailing data (degree=2)\n")
cluster.rms.eval(km.trailing.5, complete.trailing, cmodel.trailing.5.d2, prn=FALSE)
```

The degree 2 looks slightly better, though that's without any cross-validation applied.  Here's the scatterplot:
```{r cluster.scatterplot}
cluster.scatterplot.model(km.trailing.5, complete.trailing, cmodel.trailing.5.d2, sz=2)
```

This actually looks a lot better, even allowing for not holding back any data for a testing set.  For one thing, the outputs aren't confined (mostly) between +/-1 like they are in the straight regression case.  A closer look suggests that only two of the clusters, group 3 and group 4, generated models with any real predictive power.  For the other groups you see the same horizontal character displayed in the regressions over the whole dataset.  Most of the improvement in the fit for these groups comes from the intercept term.  

```{r cluster.anly}
cluster.boxplot(km.trailing.5, complete.trailing, trailing.basic)
clust.countries(complete.trailing, km.trailing.5$cluster)[c(3,4)]
```
